DRAFT \\ From physical principles to classical and quantum \\ Hamiltonian and Lagrangian particle mechanics

### 2016-07-27 2015CompleteDerivation/completeDerivation.tex
First review for relativity

The first thing to do is to amend our definition of state space to include all states at all times. It may seem like we are extending our definition of state and state space but, at a closer look, we are not. Since defined state as a physical configuration at a particular time, the set of all states (i.e. the state space) is more accurately the set of all configurations at all times. In the previous sections we tacitly simplified the problem by ignoring time, which made it easier to study the time independent case. This is common in physics and engineering but this is also the source of the above problems.

We'll therefore call \emph{complete state space} the set of all configurations at all times and we'll indicate it by $\bar{\mathcal{S}}$. In that space, a state is an element $\mathcal{s} \in \bar{\mathcal{S}}$ while a time evolution is a line $\bar{\mathcal{s}} \subset \bar{\mathcal{S}}$ that passes once at all times. Specifying an evolution map means giving a set of lines such that each state is traversed once and only once. Note that the transported state variables are perfect to label the evolutions: as they do not change during the evolution we can take them to label not just the values at a particular time, but the evolutions themselves.

With this geometric picture in mind, we can understand better the complete state space $\bar{\mathcal{C}}$ for the composite system. An evolution $\hat{\mathcal{c}}$ will be a distribution of material over all configuration at all times. Along a particle state evolution $\bar{\mathcal{s}}$, though, the distribution is constant as all the material is transported deterministically and reversibly. A state space $\mathcal{c}$, then, is a distribution over an hyper-surface that cuts across all the particle state evolutions. If we think it as a distribution over the transported state variables we are guaranteed to cover all initial conditions no matter the hyper-surface.

The first thing to do is to extend our $n$-dimensional $\mathcal{Q}$, the manifold that defines our units, to include the time variable $t$. We call $\mathcal{M}$ such extended space which is a manifold of dimension $n+1$: state variables and the time variable are independent (i.e. any combination of state and time is valid). A time variable is not simply another state variable: it doesn't identify extra states and it's the variable on which we have defined deterministic and reversible evolution. When changing variables, we still need to make sure that deterministic and reversible motion can be defined on the new time variable. Transformations like $\hat{t}=q, \; \hat{q}=t$ (exchanging time with another state variable) or $\hat{t}=\sqrt{t^2 + q^2}, \; \hat{q}=\tan^{-1}(t/q)$ (polar coordinates between time and state variable) clearly do not guarantee deterministic and reversible motion over the new time variable. To be meaningful, the change of time variable must preserve the time ordering therefore the transformation has to be strictly monotonic between the two time variables. In this sense, the time variable is special.

As before, we are interested in invariant densities which are defined on co-vectors. The co-vector $k_i dq^i$ is extended with a time component $\bar{\omega} dt$.\footnote{$\bar{\omega}$ is the classical analogue of the wave frequency. We use $\bar{\omega}$ to distinguish from the phase-space metric $\omega$.} As state variable $q^i$ have a conjugate quantity $p_i\equiv \hbar k$, the time variable $t$ will have a conjugate quantity $E\equiv\hbar\bar{\omega}$. We call the combination of $(t, E)$ the temporal degree of freedom. As time is special, so is the temporal degree of freedom, which is treated differently when defining distribution $\rho_\mathcal{c}$ and therefore by the measure $\mu$ and the metric $\omega$.

The first thing to note is that $E$ cannot be an independent variable. It cannot add any states, as those are defined by $T^*\mathcal{Q}$ only, and cannot add any time instants, as those are defined by $t$ alone. Therefore there must exist a constraint such that, locally, $E=E(t,q^i,p_i)$. The set of all states at all times is a sub-manifold of $T^*\mathcal{M}$ of dimension $2n+1$.

A composite state $\mathcal{c}$ and its distribution $\rho_\mathcal{c}$ is no longer defined at an instant of an absolute time parameter as now we are free to mix mix $t$ and $q$'s. Yet, there must be at least one time variable so that the distribution is at equal time defined over a suitable $T^*\mathcal{Q} \subset T^*\mathcal{M}$. That is, there must be at least one set of variables for which we are able to define states and their evolution. Therefore our distributions $\rho_\mathcal{c}$ are defined on $2n$-dimensional sub-manifolds of $T^*\mathcal{M}$. In other words: densities are not densities over the time variable.

Yet, as we can mix time and state variables, the volume of integration is not simply $dq^n \wedge dp_n$ as it is not invariant. Consider a change of variable $q'=q+vt$: the same surface may have projections of different sizes on $(q, p)$ than $(q', p')$. It would seem, then, that the measure for our states is different depending on the choice of variable, which cannot be.

We can understand the geometry of $T^*\mathcal{M}$ in the following way. As we have seen, the temporal degree of freedom $(t, E)$ and the other d.o.f. $(q^i, p_i)$ are not independent. So they are not orthogonal in the extended phase space. States are defined on the plane where $( q, p )$ (maximally) change. This is not the plane of constant $( t, E )$ (they are not orthogonal) where the area given by $dq \wedge dp$ is defined. It is the plane perpendicular to constant $( q, p )$. And the plane of constant $( q, p )$ is where the area given by $dt \wedge dE$ is defined. That is: the plane where we can properly count states and define our metric is perpendicular to $dt \wedge dE$.

@@

We can capture the above discussion by stating that the extended state space is a hypersurface of the symplectic manifold $(T^*\mathcal{M}, \omega)$, where $\omega \equiv \sum dq^i \wedge dp_i - dt \wedge dE$. This is the simplest manifold that allows us to define time-and-state-variable-invariant densities. The metric allows us to measure the cardinality of possibilities.

@@

\begin{prop}\label{relativistic_canonical_transformation}
	A time-dependent deterministic and reversible evolution map for a homogeneous classical material is a self-symplectomorphism on the extended phase space. That is: $\mathcal{T}_{\Delta s}: T^*\mathcal{M} \rightarrow T^*\mathcal{M}$ such that $\mathcal{T}_{\Delta s}^*\omega = \omega$ where $\mathcal{T}_{\Delta s}^*$ is the pullback of $\mathcal{T}_{\Delta s}$.
\end{prop}

@@

Note that $\mathcal{H}=0$ for all states, since $\mathcal{H}=\lambda(H-E)$ and $E = H$. For all states we also have $\lambda \neq 0$. Therefore $\mathcal{H}=0$ only because $E = H$. We can extend $\mathcal{H}$ on all of $T^* \mathcal{M}$ making sure that $\mathcal{H} \neq 0$ outside of the hyper-surface that contains states. This way we can directly use the constraint $\mathcal{H}=0$ to identify such hyper-surface.

@@

\begin{prop}\label{form_of_extended_hamiltonian}
	The form of the extended Hamiltonian is $\mathcal{H} = \lambda(t,E,q^i,p_i) \, (H(t,q^i,p_i) - E)$ where $H(t,q^i,p_i)$ is the standard Hamiltonian at each instant $t$, the conjugate time variable $E = H(t,q^i,p_i)$ represents the value of the Hamiltonian, and  $\lambda : T^*\mathcal{M} \rightarrow \mathbbm{R}$ where $\lambda(\xi^a) = d_s t$. $\mathcal{H} = 0$ for all states at all time.
\end{prop}

@@

Now that we have characterized the motion of the particles, we should also characterize the motion of the composite state.

If we have a distribution of material $\rho_\mathcal{c}(s)$ for a particular value of $s$, the first thing we know is that its support, the region where the density is non-zero, must be within the constraint $\mathcal{H} = 0$. This means $\mathcal{H} \rho_\mathcal{c}(s) = 0$ since $\mathcal{H} = 0$ where $\rho_\mathcal{c}(s) \neq 0$. If the distribution is differentiable, the evolution is given by the Poisson bracket $d_s \rho_\mathcal{c} (s) = \{ \rho_\mathcal{c}(s), \mathcal{H} \}$.

We can also consider the distribution of material at all times $\rho_\mathcal{c}\equiv \int \rho_\mathcal{c}(s) ds$. The support of the distribution is also constrained so we have $\mathcal{H} \rho_\mathcal{c} = 0$. As the density is transported over each trajectory, its value does not change over $s$. Therefore, if the distribution is differentiable, we have $d_s \rho_\mathcal{c} = \{ \rho_\mathcal{c}, \mathcal{H} \} = 0$.

\begin{prop}\label{relativistic_distribution_evolution}
	Let $\rho_\mathcal{c}(s)$ the distribution of material at a particular moment in the evolution. $\mathcal{H} \rho_\mathcal{c}(s) = 0$ and $d_s \rho_\mathcal{c} (s) = \{ \rho_\mathcal{c}(s), \mathcal{H} \}$ where $\{ f, g \} \equiv \partial_a f \omega^{ab} \partial_b g$. Let $\rho_\mathcal{c}$ the distribution of material at all times. $\mathcal{H} \rho_\mathcal{c} = 0$ and $d_s \rho_\mathcal{c} = \{ \rho_\mathcal{c}, \mathcal{H} \} = 0$.
\end{prop}

\begin{justification}
	We claim $\mathcal{H} \rho_\mathcal{c}(s) = 0 \; \forall \mathcal{c} \in \mathcal{C} \wedge s \in \mathbbm{R}$. Let $\rho_\mathcal{c}(s) : T^*\mathcal{M} \rightarrow \mathbbm{R}$ be the distribution associated with a composite state $\mathcal{c}$ at a given value of $s$. The distribution is defined on physical states. By \ref{form_of_extended_hamiltonian} $supp(\rho_\mathcal{c}(s)) \subseteq T^*\mathcal{M} \,\backslash\, supp(\mathcal{H})$. $supp(\mathcal{H} \rho_\mathcal{c}(s)) = supp(\mathcal{H}) \cap supp(\rho_\mathcal{c}(s)) \subseteq \varnothing$.

@@

As mentioned at the beginning of the section, no new assumption was required. Relativistic Hamiltonian mechanics is simply the correct way of handling time dependent motion. As such, there was no mention of the speed of light $c$, $\mathcal{Q}$ being a Riemannian manifold or the metric $g$. These will be introduced in the next chapter with the kinematic equivalence assumption. In light of this, it is probably better to consider relativistic mechanics as a feature of the geometry of the extended phase space more than the geometry of space-time.

@@

>>>>

The first thing to do is to amend our definition of state space to include all states at all times. It may seem like we are extending our definition of state and state space but, at a closer look, we are not. Since we defined state as a physical configuration at a particular time, the set of all states (i.e. the state space) is more accurately defined as the set of all configurations at all times. In the previous sections we tacitly simplified the problem by ignoring time, which made it easier to study the time independent case. This is common in physics and engineering, which is why we kept the standard name, but this is also the source of the above problems.

We'll therefore call \emph{complete state space} of particle states the set of all configurations at all times and we'll indicate it by $\bar{\mathcal{S}}$. In that space, a state is an element $\mathcal{s} \in \bar{\mathcal{S}}$ while a time evolution is a line $\bar{\mathcal{s}} \subset \bar{\mathcal{S}}$ that passes once at all times. Specifying an evolution map means giving a set of lines such that each state is traversed once and only once. Locally, each state will be given by $(q^i, k_i, t)$ therefore $\bar{\mathcal{S}}$ is a $2n+1$ dimensional manifold. A region at equal time will be a hyper-surface $\mathcal{S}_{t=t_0}$ that includes all possible configuration at a particular time. Such hyper-surface will need to cut across all evolutions once and only once, therefore not all functions of $\bar{\mathcal{S}}$ are suitable time variables.

\begin{prop}
	The complete state space $\bar{\mathcal{S}}$ for the particles of a classical material is a $2n+1$ dimensional differentiable manifold. The state space $\mathcal{S}_{t=t_0}$ at a particular time is a hyper-surface of $\bar{\mathcal{S}}$.
\end{prop}

\begin{justification}
	TODO
\end{justification}

For the composite state we need to take a different approach. A distribution of material over states at equal time $\rho_\mathcal{c} : \mathcal{S}_{t=t_0} \rightarrow \mathbbm{R}$ will depend on the choice of time variable, therefore the complete state space will not be useful as it will not be variable independent. What we are really interested in is the \emph{evolution space} of a classical material, which we'll indicated by $\bar{\mathcal{C}}$. To each evolution $\bar{\mathcal{c}} \in \bar{\mathcal{C}}$ will correspond a distribution $\rho_{\bar{\mathcal{c}}} : \bar{\mathcal{S}} \rightarrow \mathbbm{R}$ over all particle states at all times. This distribution will be differentiable, it's differentiable over all equal time hyper-surfaces and time evolution is differentiable, but is not integrable over the complete state space $\bar{\mathcal{S}}$. Such integration will operate over time as well, giving the amount of material multiplied by time. If time is taken to be infinite the value is not finite. The distribution, instead, must be integrable over all equal time hyper-surfaces: no matter what time variable we choose, the amount of material found at a particular time has to be finite.

\begin{prop}
	The evolution state space $\bar{\mathcal{C}}$ for a classical material is isomorphic to the space of differentiable functions that are Lebesgue integrable on any equal time hyper-surface. That is $\bar{\mathcal{C}} \cong  C^1(\bar{\mathcal{S}})\cap L^1_t(\bar{\mathcal{S}}, \mu)$ where $L^1_t(\bar{\mathcal{S}}, \mu)=\{{\rho: \bar{\mathcal{S}} \rightarrow \mathbbm{R}} \; | \; {\int_{\mathcal{S}_{t=t_0}} |\rho| d\mu < \infty} \; \forall \, t, t_0 \}$.
\end{prop}

\begin{justification}
\end{justification}

Before proceeding, it's useful to get a better intuition for the geometry of these space. Note that the transported state variables are perfect to label the particle state evolution evolutions $\bar{\mathcal{s}} \subset \bar{\mathcal{S}}$: as they do not change during the evolution we can take them to label not just the values at a particular time, but the evolutions themselves. The distribution is transported deterministically and reversibly over those lines, so the value of the distribution will be constant along each particle state evolution $\bar{\mathcal{s}}$. To identify the full evolution, then, we just need to specify the distribution only on a region that cuts across all particle evolution, i.e. a hyper-surface at constant time.

\subsection{Relativistic cotangent bundle}

We now want to find the metric for our complete state space $\bar{\mathcal{S}}$. The first thing to do is to complete our $n$-dimensional $\mathcal{Q}$, the manifold that defines our units, to include the time variable $t$. In fact, the idea that the set of units to identify a physical configuration at a particular time did not include time should look, retrospectively, shaky at best. We call $\mathcal{M}$ such space which is a manifold of dimension $n+1$: state variables and the time variable are independent (i.e. any combination of state and time is valid). As we said before, the time variable is not simply another state variable: it doesn't identify extra states and it's the variable on which we have defined deterministic and reversible evolution. When changing variables, we still need to make sure that deterministic and reversible motion can be defined on the new time variable. Transformations like $\hat{t}=q, \; \hat{q}=t$ (exchanging time with another state variable) or $\hat{t}=\sqrt{t^2 + q^2}, \; \hat{q}=\tan^{-1}(t/q)$ (polar coordinates between time and state variable) clearly do not guarantee deterministic and reversible motion over the new time variable. To be meaningful, the change of time variable must preserve time ordering therefore the transformation has to be strictly monotonic between the two time variables. On the other hand, the change of state variables must preserve the ability to uniquely identify states each instant in time. In this sense, time is not just like another state variable.\footnote{In relativity, for example, it is common to use foliations and the $3+1$ formalism to recover the special character of time when performing calculations or to gain better physical insight.}

As in the time independent case, we are interested in invariant densities which are defined on co-vectors. The co-vector $k_i dq^i$ is completed with a time component $\bar{\omega} dt$.\footnote{$\bar{\omega}$ is the classical analogue of the wave frequency. We use $\bar{\omega}$ to distinguish from the phase-space metric $\omega$.} As state variable $q^i$ have a conjugate quantity $p_i\equiv \hbar k$, the time variable $t$ will have a conjugate quantity $E\equiv\hbar\bar{\omega}$. We call the combination of $(t, E)$ the temporal degree of freedom. As time is special, so is the temporal degree of freedom, which is treated differently when defining distribution $\rho_\mathcal{c}$ and therefore by the measure $\mu$ and the metric $\omega$.

Most of all, the temporal degree of freedom is not an independent degree of freedom because $E$ cannot be an independent variable. It cannot add any states at a particular time, as those are defined by $T^*\mathcal{Q}$ only, and cannot add any time instants, as those are defined by $t$ alone. Therefore there must exist a constraint such that, locally, $E=E(t,q^i,p_i)$. The complete state space set $\bar{\mathcal{S}}$ is a hyper-surface of $T^*\mathcal{M}$.

As the temporal degree of freedom $(t, E)$ is not independent of the other d.o.f. $(q^i, p_i)$, it is not orthogonal to them in $T^*\mathcal{M}$. States are defined on the plane where $( q, p )$ (maximally) change. This is not the plane of constant $( t, E )$ (they are not orthogonal) where the area given by $dq \wedge dp$ is defined. It is the plane perpendicular to constant $( q, p )$. And the plane of constant $( q, p )$ is where the area given by $dt \wedge dE$ is defined. That is: the plane where we can properly count states and define our metric is perpendicular to $dt \wedge dE$.

@@

We can capture the above discussion by stating that the complete state space $\bar{\mathcal{S}}$ is a hypersurface of the symplectic manifold $(T^*\mathcal{M}, \omega)$, where $\omega \equiv \sum dq^i \wedge dp_i - dt \wedge dE$. This is the manifold that allows us to define time-and-state-variable-invariant densities. The metric allows us to measure the cardinality of possibilities.

@@

\begin{prop}\label{relativistic_canonical_transformation}
	A time-dependent deterministic and reversible evolution map for a homogeneous classical material is a self-symplectomorphism on $T^*\mathcal{M}$. That is: $\mathcal{T}_{\Delta s}: T^*\mathcal{M} \rightarrow T^*\mathcal{M}$ such that $\mathcal{T}_{\Delta s}^*\omega = \omega$ where $\mathcal{T}_{\Delta s}^*$ is the pullback of $\mathcal{T}_{\Delta s}$.
\end{prop}

@@

Note that $\mathcal{H}=0$ for all states $\mathcal{s} \in \bar{\mathcal{S}} \subset T^* \mathcal{M}$, since $\mathcal{H}=\lambda(H-E)$ and $E = H$. For all states we also have $\lambda \neq 0$. Therefore $\mathcal{H}=0$ only because $E = H$. We can extend $\mathcal{H}$ on all of $T^* \mathcal{M}$ making sure that $\mathcal{H} \neq 0$ outside of the complete state space $\bar{\mathcal{S}}$. This way we can directly use the constraint $\mathcal{H}=0$ to identify $\bar{\mathcal{S}}$ as a subspace of $T^* \mathcal{M}$.

@@

\begin{prop}\label{form_of_extended_hamiltonian}
	The form of the extended Hamiltonian is $\mathcal{H} = \lambda(t,E,q^i,p_i) \, (H(t,q^i,p_i) - E)$ where $H(t,q^i,p_i)$ is the standard Hamiltonian at each instant $t$, the conjugate time variable $E = H(t,q^i,p_i)$ represents the value of the Hamiltonian, and  $\lambda : T^*\mathcal{M} \rightarrow \mathbbm{R}$ where $\lambda(\xi^a) = d_s t$. $\bar{\mathcal{S}}$ is the level set for $\mathcal{H} = 0$.
\end{prop}

@@

Now that we have characterized the motion of the particles, we should also characterize the evolution of composite states.

Suppose we have a distribution of material $\rho_{\bar{\mathcal{c}}}(s)$ associated with an evolution $\bar{\mathcal{c}}$. Its support, the region where the density is non-zero, is within $\bar{\mathcal{S}}$. Therefore  $\mathcal{H} \rho_{\bar{\mathcal{c}}}(s) = 0$ since $\mathcal{H} = 0$ in $\bar{\mathcal{S}}$ where $\rho_{\bar{\mathcal{c}}}(s) \neq 0$. As the density is transported over each trajectory, its value does not change over $s$. Therefore we have $d_s \rho_{\bar{\mathcal{c}}} = \{ \rho_{\bar{\mathcal{c}}}, \mathcal{H} \} = 0$.

\begin{prop}\label{relativistic_distribution_evolution}
	Let $\rho_{\bar{\mathcal{c}}}(s)$ the distribution of material associated with the evolution $\bar{\mathcal{c}}$. $\mathcal{H} \rho_{\bar{\mathcal{c}}} = 0$ and $d_s \rho_{\bar{\mathcal{c}}} = \{ \rho_{\bar{\mathcal{c}}}, \mathcal{H} \} = 0$.
\end{prop}

\begin{justification}
	TODO: language consistency

	We claim $\mathcal{H} \rho_\mathcal{c}(s) = 0 \; \forall \mathcal{c} \in \mathcal{C} \wedge s \in \mathbbm{R}$. Let $\rho_\mathcal{c}(s) : T^*\mathcal{M} \rightarrow \mathbbm{R}$ be the distribution associated with a composite state $\mathcal{c}$ at a given value of $s$. The distribution is defined on physical states. By \ref{form_of_extended_hamiltonian} $supp(\rho_\mathcal{c}(s)) \subseteq T^*\mathcal{M} \,\backslash\, supp(\mathcal{H})$. $supp(\mathcal{H} \rho_\mathcal{c}(s)) = supp(\mathcal{H}) \cap supp(\rho_\mathcal{c}(s)) \subseteq \varnothing$.

@@

As mentioned at the beginning of the section, no new assumption was required. Relativistic Hamiltonian mechanics is simply the correct way of handling time dependent motion. As such, there was no mention of the speed of light $c$, $\mathcal{M}$ being a pseudo-Riemannian manifold or its metric $g$. These will be introduced in the next chapter with the kinematic equivalence assumption. In light of this, it is probably better to consider relativistic mechanics as a feature of the geometry of the complete state space more than the geometry of space-time. Also note that this feature arises from the use of the standard topology on $\mathbbm{R}^n$ and the need of invariant densities. It would not have arisen with the use of a discrete topology.

@@

As a final thought, we note how the physics maps less elegantly to the math in the time dependent case. The function space for $\bar{\mathcal{C}}$ is not simply the Lebesgue integrable differentiable functions, but those that are integrable on particular hyper-surfaces. The complete state space $\bar{\mathcal{S}}$ is a hyper-surface of $T^*\mathcal{M}$. The extended Hamiltonian is needed to identify that subspace. This may be a hint that a better formulation is lurking. One where we consider the particle evolutions $\bar{\mathcal{s}}$ as the primary objects, label those (instead of the states). The distribution of material would simply be a distribution over the particle evolutions, a Lebesgue integrable differentiable functions.
